---
title: "Data Science for Public Policy"
subtitle: "Applied Introduction to R's Tidyverse"
author: "Lezhi Charles Cao - lc1267"
format: 
  pdf:
    code-line-numbers: true
execute: 
  warning: false
urlcolor: blue
---

\vspace{0.1in}

```{r}
library(tidyverse)
library(ipumsr)
library(srvyr)
library(haven)
library(ggplot2)
library(treemapify)
library(ggridges)

```


# Visualization 01 (1 point)

This treemap visualizes the cumulative gun deaths by county for New York (NY), Pennsylvania (PA), and New Jersey (NJ) from 1999 to 2019. It highlights the disparities in gun deaths across counties within these states, with counties having fewer than 100 cumulative deaths combined into the "Other" category. This information is crucial for policymakers, as it identifies regions with higher gun violence rates, allowing for targeted intervention and resource allocation.

```{r}
gun_d <- read_csv("gun_deaths_us_1999_2019.csv")

gun.3states <- filter(
  gun_d, 
  (State == "NY") | (State == "PA") | (State == "NJ")
  ) 
gun.3states.d <- gun.3states %>% 
  group_by(State, County) %>% 
  summarize(total_deaths = sum(Deaths), 
            .groups = "drop") 

gun.3states.d <- gun.3states.d %>% 
  group_by(State, County) %>% 
  summarise(total_deaths = sum(total_deaths)) %>% 
  ungroup() %>% 
  mutate(County = if_else(total_deaths < 100, 
                          "Other", County))  %>%
  group_by(State, County) 
#counties with total gun death lower than 100 is combined into the
#other category

ggplot(gun.3states.d, aes(
  area = total_deaths, 
  fill = State, 
  label = County, 
  subgroup = State)) + 
  geom_treemap(show.legend = FALSE)+ 
  geom_treemap_subgroup_border(
    color = "white", 
    show.legend = FALSE)+ 
  geom_treemap_text(
    fontface = "italic", 
    color = "white", 
    place = "center",  
    min.size = 0.1, 
    reflow = T)+ 
  geom_treemap_subgroup_text(
    place = "center", 
    alpha = 0.3, 
    color = "lightgrey", 
    min.size =0,
    grow = T)+ 
  scale_fill_manual(
    values = c("NY" = "skyblue3", 
               "PA" = "#002147",
               "NJ" = "darkgoldenrod"))+ 
  #color chosen from the different state colors 
  labs(title= (paste ("Cumulative Gun Deaths by County for New York (NY), Pennsylvania (PA) and \nNew Jersey (NJ): 1999 - 2019*")), 
       subtitle = "PA has the highest cumulative Gun Deaths by State and Individual County", 
       caption = "*Counties with cumulative gun deaths lower than 100 are merged into the other category
Data: Centers for Disease Control and Prevention") +
  theme(plot.title = element_text(size=11, face = "bold"),
        plot.subtitle = element_text(size=8.5, face = "italic"),
        plot.caption = element_text(size = 7))

```

Data URL: https://data.world/nkrishnaswami/gun-deaths-by-county/workspace/project-summary?agentid=nkrishnaswami&datasetid=gun-deaths-by-county
\newpage

# Visualization 02 (2 points)

This graph compares the population and gun deaths for New Jersey (NJ), New York (NY), and Pennsylvania (PA) from 1999 to 2019.It demonstrates that while NJ and NY exhibit similar population and gun death trends, PA's gun death trend begins to deviate from population growth since 2014. This insight is vital for policymakers to understand the dynamics between population changes and gun violence and develop effective strategies for reducing gun deaths.

```{r}
gun.states <- 
  gun_d %>%
  filter(State == "PA" | State == "NY" | State == "NJ") %>%
  group_by(State,Year) %>%
  select(State,Year,Population,Deaths) %>%
  summarize(deaths=sum(Deaths), population = sum(Population)) 

ggplot(gun.states, aes(x = Year)) +
  geom_line(aes(y = deaths, 
                color = "Gun Deaths"), 
            size = 2) +
  geom_line(aes(y = population/11000, color = "Population"), size = 1) +
  scale_color_manual(
    values = c("Population" = "lightblue", "Gun Deaths" = "pink"),
    labels = c("Gun Deaths", "Population")
  ) +
  facet_wrap(~ State, ncol = 3) +
  labs(
    title = "Population and Gun Deaths for New Jersey (NJ), New Jersey (NJ) and \nPennslyvania (PA): 1999 - 2019",
    subtitle = "On average, NJ and NY's population and gun death have a similar trend, while PA's gun death trend \nstart to deviate from population trend since 2014.",
    caption = "Data: Centers for Disease Control and Prevention") +
  theme_minimal() +
  scale_y_continuous(
    name = "Gun Deaths",
    limits = c(0, 1600),
    breaks = seq(0, 1600, 100),
    sec.axis = sec_axis(
      ~ . *11000,
      name = "Population",
      breaks = seq(5*10^6, 17600000, by = 1100000),
      labels = c("5M", "6.1M", "7.2M", "8.3M", "9.4M", "10.5M", "11.6M","12.7M", "13.8M", "14.9M", "16M", "17.1M")
    )
  )+  
   scale_x_continuous(
    breaks = seq(1999, 2019, 5), labels = seq(1999, 2019, 5)   
  )+
  theme(plot.title = element_text(size=11, face = "bold"),
        plot.subtitle = element_text(size=8.5, face = "italic"),
        plot.caption = element_text(size = 6.5),
        legend.title = element_blank(), 
        legend.text = element_text(size = 6),
        legend.position = "bottom",
        axis.text = element_text(size=7),
        axis.title = element_text(size=7),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())

#This graph compares the population and gun deaths for New Jersey (NJ), New York (NY), and Pennsylvania (PA) from 1999 to 2019.
#It demonstrates that while NJ and NY exhibit similar population and gun death trends, PA's gun death trend begins to deviate from population growth since 2014.
#This insight is vital for policymakers to understand the dynamics between population changes and gun violence. In particular, why nearby state like PA showed different trends between gun deaths and population. This visualization could serve as a starting point for policymakers to develop


```
\newpage

# Visualization 03 (2 points)

This bar chart compares total gun deaths by county for the selected years 1999, 2009, and 2019, in NY, PA, and NJ. It provides a clear visualization of the variations in gun deaths over time and between counties within these states. Policymakers can use this information to identify counties with the highest gun death rates and implement localized policies to address this issue effectively.

```{r}
# Filter the data for selected years and states
selected_years <- c(1999, 2009, 2019)
filtered_data <- gun_d %>%
  filter(State %in% c("NJ"), Year %in% selected_years, Deaths < 30) 

# Bar Chart Comparing Counties for Selected States (1999, 2009, 2019)
ggplot(filtered_data, aes(x = reorder(County, -Deaths), y = Deaths)) +
  geom_bar(stat = "identity", fill = "darkgoldenrod") +
  labs(
    title = "New Jersey's Counties with Total Gun Deaths under 30: 1999, 2009, 2019",
    subtitle = "Counties like Middlesex saw a decrease in Gun Death then a increase as time passes",
    caption = "Data: Centers for Disease Control and Prevention" 
   ) +
  facet_wrap(~ Year, ncol = 3) +
  theme_minimal() +
  ylab("Cumulative Gun Deaths")+
  theme(
    plot.title = element_text(size = 9.5, face = "bold"),
    plot.subtitle = element_text(size = 8, face = "italic"),
    plot.caption = element_text(size = 6.5),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 8),
    axis.text.x = element_text(size = 5, angle = 45, hjust = 1)
  ) 
```
\newpage

# Exercise 04 (2 points)

This pie chart illustrates the distribution of gun deaths in 5-year intervals from 1999 to 2019 for NY, PA, and NJ. 
It allows policymakers to see how gun deaths are distributed over time, which can inform long-term policy planning and highlight any significant shifts in gun violence patterns.

```{r}
# Calculate 5-year intervals for the years 1999-2019
intervals <- seq(1999, 2019, by = 5)

# Create a dataset with the total gun deaths in each 5-year interval
gun_deaths_intervals <- gun_d %>%
  filter(State %in% c("NY", "PA", "NJ")) %>%
  mutate(Interval = cut(Year, breaks = intervals, labels = intervals[-1], include.lowest = TRUE)) %>%
  group_by(State, Interval) %>%
  summarize(TotalDeaths = sum(Deaths))

  ggplot(gun_deaths_intervals, aes(x = "", y = TotalDeaths, fill = as.factor(Interval))) +
  geom_bar(stat = "identity") +
  coord_polar(theta = "y") +
  labs(
    title = "Distribution of Gun Deaths by 5-Year Intervals (1999-2019)",
    subtitle = "The Proportion of Gun Deaths in 5-Year Intervals",
    caption = "Data: Centers for Disease Control and Prevention"
  ) +
  scale_fill_discrete(name = "5-Year Interval") +
  theme_void() +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    plot.subtitle = element_text(size = 8.5),
    plot.caption = element_text(size = 6.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 6)
  )
```

